A Plot is Worth a Thousand Tests: Assessing Residual Diagnostics with the Lineup Protocol

### 2023-04-22 paper.Rmd
up to section 3

Figure \ref{fig:residual-plot-common-departures}D) shows a scatterplot where the residuals have a skewed distribution, as seen by the uneven vertical spread. Unlike non-linearity and heteroskedasticity, non-normality is usually detected with a different type of residual plot: a histogram or normal probability plot.  Because we focus on scatterplots, non-normality is not one of the departures examined in depth in this paper. (@loy_variations_2016 and @loy2013diagnostic discuss related work on non-normality checking.)

@@

The examination of data plots to infer signal or patterns (or lack thereof) is fraught with variation in the human ability to interpret and decode the information embedded in a graph [@cleveland_graphical_1984]. To rely on human examination thus feels as though there is a subjective nature to using diagnostic plots to infer appropriateness of statistical models. Ideally we would expect visual evaluation should be done objectively and human-to-human variation in reading data plots small, but it cannot be assured.

In practice, over-interpretation of a single plot is common.  For instance, @roy_chowdhury_using_2015, described a published example where the authors over-interpreted separation between gene groups from a two-dimensional projection of a linear discriminant analysis. @roy_chowdhury_using_2015 showed that there were no differences in the expression levels between the gene groups. The lineup protocol proposed in @buja_statistical_2009 provides a solution to over-interpretation.  @majumder_validation_2013 showed that its use is analogous to the null hypothesis significance testing (NHST) framework. More specifically, the protocol consists of $m$ randomly placed plots, where one plot is the data plot, and the remaining $m - 1$ plots, referred to as the _null plots_, have the identical graphical procedure as the data plot except the data is replaced with a data generation mechanism that is consistent with the null hypothesis, $H_0$. Then, an observer who have not seen the data plot will be asked to point out the most different plot from the lineup. Under $H_0$, it is expected that the data plot would have no distinguishable difference from the null plots, and the probability that the observer correctly picks the data plot is $1/m$. If one rejects $H_0$ as the observer correctly picks the data plot, then the Type I error of this test is $1/m$. This protocol requires apriori specification of $H_0$ (or at least a null data generating mechanism) much like the requirement of knowing the distribution of the test statistic in NHST.

@@

Data used in the $m - 1$ null plots needs to be simulated. In regression diagnostics, sampling data consistent with $H_0$ is equivalent to sampling data from the assumed model. As @buja_statistical_2009 suggested, $H_0$ is usually a composite hypothesis controlled by nuisance parameters. Since regression models can have various forms, there is no general solution to this problem, but it sometimes can be reduced to a so called "reference distribution" by applying one of the three methods: (i) sampling from a conditional distribution given a minimal sufficient statistic under $H_0$, (ii) parametric bootstrap sampling with nuisance parameters estimated under $H_0$, and (iii) Bayesian posterior predictive sampling. The conditional distribution given a minimal sufficient statistic is the best justified reference distribution among the three [@buja_statistical_2009]. Essentially, null residuals can be simulated by regressing $N$ i.i.d standard normal random draws on the predictors, then rescaling it by the ratio of residual sum of square in two regressions.

The effectiveness of lineup protocol for regression analysis is validated by @majumder_validation_2013 under relatively simple settings with up to two predictors. Their results suggest that visual tests are capable of testing the significance of a single predictor with a similar power as a t-test, though they express that in general it is unnecessary to use visual inference if there exists a conventional test, and they do not expect the visual test to perform equally well as the conventional test. In their third experiment, where there is not a conventional test, visual test outperforms the conventional test for a large margin. This is encouraging, as it promotes the use of visual inference in situations where there are no existing statistical testing procedures. Visual inference have also been integrated into diagnostic of hierarchical linear models by @loy2013diagnostic, @loy2014hlmdiag and @loy2015you. They use lineup protocols to judge the assumption of linearity, normality and constant error variance for both the level-1 and level-2 residuals.

>>>>

Figure \ref{fig:residual-plot-common-departures}D shows a scatterplot where the residuals have a skewed distribution, as seen by the uneven vertical spread. Unlike non-linearity and heteroskedasticity, non-normality is usually detected with a different type of residual plot: a histogram or normal probability plot.  Because we focus on scatterplots, non-normality is not one of the departures examined in depth in this paper. (loy_variations_2016 discuss related work on non-normality checking.)

@@

The examination of data plots to infer signal or patterns (or lack thereof) is fraught with variation in the human ability to interpret and decode the information embedded in a graph [@cleveland_graphical_1984]. Human examination of diagnostic plots can feel subjective. Ideally we would expect visual evaluation to be done objectively with small amounts of human-to-human variation in reading data plots, but we cannot assume this is the case.

In practice, over-interpretation of a single plot is common.  For instance, @roy_chowdhury_using_2015 described a published example where authors over-interpreted separation between gene groups from a two-dimensional projection of a linear discriminant analysis. @roy_chowdhury_using_2015 showed that there were no differences in the expression levels between the gene groups. One solution to over-interpretation is to examine the plot in the context of natural sampling variability assumed by the model, called the lineup protocol, as proposed in @buja_statistical_2009.  @majumder_validation_2013 showed that the lineup protocol is analogous to the null hypothesis significance testing framework. The protocol consists of $m$ randomly placed plots, where one plot is the data plot, and the remaining $m - 1$ plots, referred to as the _null plots_, are constructed using the same graphical procedure as the data plot but the data is replaced with null data that is generated in a manner consistent with the null hypothesis, $H_0$. Then, an observer who has not seen the data plot is asked to point out the most different plot from the lineup. Under $H_0$, it is expected that the data plot would have no distinguishable difference from the null plots, and the probability that the observer correctly picks the data plot is $1/m$. If one rejects $H_0$ as the observer correctly picks the data plot, then the Type I error of this test is $1/m$. This protocol requires a priori specification of $H_0$ (or at least a null data generating mechanism), much like the requirement of knowing the sampling distribution of the test statistic in null hypothesis significance testing framework.

@@

Data used in the $m - 1$ null plots is typically simulated. In regression diagnostics, sampling data consistent with $H_0$ is equivalent to sampling data from the assumed model. As @buja_statistical_2009 suggested, $H_0$ is usually a composite hypothesis controlled by nuisance parameters. Since regression models can have various forms, there is no general solution to this problem, but it sometimes can be reduced to a so called "reference distribution" by applying one of the three methods: (i) sampling from a conditional distribution given a minimal sufficient statistic under $H_0$, (ii) parametric bootstrap sampling with nuisance parameters estimated under $H_0$, and (iii) Bayesian posterior predictive sampling. The conditional distribution given a minimal sufficient statistic is the best justified reference distribution among the three [@buja_statistical_2009]. Essentially, null residuals can be simulated by regressing $N$ i.i.d standard normal random draws on the predictors, then rescaling it by the ratio of residual sum of square in two regressions.

The effectiveness of lineup protocol for regression analysis has been validated by @majumder_validation_2013 under relatively simple settings with up to two predictors. Their results suggest that visual tests are capable of testing the significance of a single predictor with a similar power to a t-test, though they express that in general it is unnecessary to use visual inference if there exists a conventional test, and they do not expect the visual test to perform equally well as the conventional test. In their third experiment, where there is not a conventional test, visual test outperforms the conventional test by a large margin. This is encouraging, as it supports the use of visual inference in situations where there are no existing statistical testing procedures. Visual inference has also been integrated into diagnostics for hierarchical linear models where the lineup protocol is used to judge the assumptions of linearity, normality and constant error variance for both the level-1 and level-2 residuals [@loy2013diagnostic, @loy2014hlmdiag and @loy2015you].

### 2023-04-22 paper.tex
up to section 3

Figure \ref{fig:residual-plot-common-departures}D) shows a scatterplot
where the residuals have a skewed distribution, as seen by the uneven
vertical spread. Unlike non-linearity and heteroskedasticity,
non-normality is usually detected with a different type of residual
plot: a histogram or normal probability plot. Because we focus on
scatterplots, non-normality is not one of the departures examined in
depth in this paper. (\citet{loy_variations_2016} and
\citet{loy2013diagnostic} discuss related work on non-normality
checking.)

@@

The examination of data plots to infer signal or patterns (or lack
thereof) is fraught with variation in the human ability to interpret and
decode the information embedded in a graph
\citep{cleveland_graphical_1984}. To rely on human examination thus
feels as though there is a subjective nature to using diagnostic plots
to infer appropriateness of statistical models. Ideally we would expect
visual evaluation should be done objectively and human-to-human
variation in reading data plots small, but it cannot be assured.

In practice, over-interpretation of a single plot is common. For
instance, \citet{roy_chowdhury_using_2015}, described a published
example where the authors over-interpreted separation between gene
groups from a two-dimensional projection of a linear discriminant
analysis. \citet{roy_chowdhury_using_2015} showed that there were no
differences in the expression levels between the gene groups. The lineup
protocol proposed in \citet{buja_statistical_2009} provides a solution
to over-interpretation. \citet{majumder_validation_2013} showed that its
use is analogous to the null hypothesis significance testing (NHST)
framework. More specifically, the protocol consists of \(m\) randomly
placed plots, where one plot is the data plot, and the remaining
\(m - 1\) plots, referred to as the \emph{null plots}, have the
identical graphical procedure as the data plot except the data is
replaced with a data generation mechanism that is consistent with the
null hypothesis, \(H_0\). Then, an observer who have not seen the data
plot will be asked to point out the most different plot from the lineup.
Under \(H_0\), it is expected that the data plot would have no
distinguishable difference from the null plots, and the probability that
the observer correctly picks the data plot is \(1/m\). If one rejects
\(H_0\) as the observer correctly picks the data plot, then the Type I
error of this test is \(1/m\). This protocol requires apriori
specification of \(H_0\) (or at least a null data generating mechanism)
much like the requirement of knowing the distribution of the test
statistic in NHST.

@@

Data used in the \(m - 1\) null plots needs to be simulated. In
regression diagnostics, sampling data consistent with \(H_0\) is
equivalent to sampling data from the assumed model. As
\citet{buja_statistical_2009} suggested, \(H_0\) is usually a composite
hypothesis controlled by nuisance parameters. Since regression models
can have various forms, there is no general solution to this problem,
but it sometimes can be reduced to a so called ``reference
distribution'' by applying one of the three methods: (i) sampling from a
conditional distribution given a minimal sufficient statistic under
\(H_0\), (ii) parametric bootstrap sampling with nuisance parameters
estimated under \(H_0\), and (iii) Bayesian posterior predictive
sampling. The conditional distribution given a minimal sufficient
statistic is the best justified reference distribution among the three
\citep{buja_statistical_2009}. Essentially, null residuals can be
simulated by regressing \(N\) i.i.d standard normal random draws on the
predictors, then rescaling it by the ratio of residual sum of square in
two regressions.

The effectiveness of lineup protocol for regression analysis is
validated by \citet{majumder_validation_2013} under relatively simple
settings with up to two predictors. Their results suggest that visual
tests are capable of testing the significance of a single predictor with
a similar power as a t-test, though they express that in general it is
unnecessary to use visual inference if there exists a conventional test,
and they do not expect the visual test to perform equally well as the
conventional test. In their third experiment, where there is not a
conventional test, visual test outperforms the conventional test for a
large margin. This is encouraging, as it promotes the use of visual
inference in situations where there are no existing statistical testing
procedures. Visual inference have also been integrated into diagnostic
of hierarchical linear models by \citet{loy2013diagnostic},
\citet{loy2014hlmdiag} and \citet{loy2015you}. They use lineup protocols
to judge the assumption of linearity, normality and constant error
variance for both the level-1 and level-2 residuals.

>>>>

Figure \ref{fig:residual-plot-common-departures}D shows a scatterplot
where the residuals have a skewed distribution, as seen by the uneven
vertical spread. Unlike non-linearity and heteroskedasticity,
non-normality is usually detected with a different type of residual
plot: a histogram or normal probability plot. Because we focus on
scatterplots, non-normality is not one of the departures examined in
depth in this paper. (loy\_variations\_2016 discuss related work on
non-normality checking.)

@@

The examination of data plots to infer signal or patterns (or lack
thereof) is fraught with variation in the human ability to interpret and
decode the information embedded in a graph
\citep{cleveland_graphical_1984}. Human examination of diagnostic plots
can feel subjective. Ideally we would expect visual evaluation to be
done objectively with small amounts of human-to-human variation in
reading data plots, but we cannot assume this is the case.

In practice, over-interpretation of a single plot is common. For
instance, \citet{roy_chowdhury_using_2015} described a published example
where authors over-interpreted separation between gene groups from a
two-dimensional projection of a linear discriminant analysis.
\citet{roy_chowdhury_using_2015} showed that there were no differences
in the expression levels between the gene groups. One solution to
over-interpretation is to examine the plot in the context of natural
sampling variability assumed by the model, called the lineup protocol,
as proposed in \citet{buja_statistical_2009}.
\citet{majumder_validation_2013} showed that the lineup protocol is
analogous to the null hypothesis significance testing framework. The
protocol consists of \(m\) randomly placed plots, where one plot is the
data plot, and the remaining \(m - 1\) plots, referred to as the
\emph{null plots}, are constructed using the same graphical procedure as
the data plot but the data is replaced with null data that is generated
in a manner consistent with the null hypothesis, \(H_0\). Then, an
observer who has not seen the data plot is asked to point out the most
different plot from the lineup. Under \(H_0\), it is expected that the
data plot would have no distinguishable difference from the null plots,
and the probability that the observer correctly picks the data plot is
\(1/m\). If one rejects \(H_0\) as the observer correctly picks the data
plot, then the Type I error of this test is \(1/m\). This protocol
requires a priori specification of \(H_0\) (or at least a null data
generating mechanism), much like the requirement of knowing the sampling
distribution of the test statistic in null hypothesis significance
testing framework.

@@

Data used in the \(m - 1\) null plots is typically simulated. In
regression diagnostics, sampling data consistent with \(H_0\) is
equivalent to sampling data from the assumed model. As
\citet{buja_statistical_2009} suggested, \(H_0\) is usually a composite
hypothesis controlled by nuisance parameters. Since regression models
can have various forms, there is no general solution to this problem,
but it sometimes can be reduced to a so called ``reference
distribution'' by applying one of the three methods: (i) sampling from a
conditional distribution given a minimal sufficient statistic under
\(H_0\), (ii) parametric bootstrap sampling with nuisance parameters
estimated under \(H_0\), and (iii) Bayesian posterior predictive
sampling. The conditional distribution given a minimal sufficient
statistic is the best justified reference distribution among the three
\citep{buja_statistical_2009}. Essentially, null residuals can be
simulated by regressing \(N\) i.i.d standard normal random draws on the
predictors, then rescaling it by the ratio of residual sum of square in
two regressions.

The effectiveness of lineup protocol for regression analysis has been
validated by \citet{majumder_validation_2013} under relatively simple
settings with up to two predictors. Their results suggest that visual
tests are capable of testing the significance of a single predictor with
a similar power to a t-test, though they express that in general it is
unnecessary to use visual inference if there exists a conventional test,
and they do not expect the visual test to perform equally well as the
conventional test. In their third experiment, where there is not a
conventional test, visual test outperforms the conventional test by a
large margin. This is encouraging, as it supports the use of visual
inference in situations where there are no existing statistical testing
procedures. Visual inference has also been integrated into diagnostics
for hierarchical linear models where the lineup protocol is used to
judge the assumptions of linearity, normality and constant error
variance for both the level-1 and level-2 residuals
\citep[\citet{loy2014hlmdiag} and
\citet{loy2015you}]{loy2013diagnostic}.
